I am grateful to David Hogg for being a great advisor and teacher. 
I thank him for his consistent support, encouragement, patience, teachings, 
and all the valuable skills I learned from him.

I thank David Hogg, Jeremy Tinker, Michael Blanton, Kyle Cranmer, Roman Scoccimarro, and Boris Leistedt for 
reading an initial draft of thesis and sharing their valuable comments with me.

Throughout my PhD years, I have had the opportunity to immensly benefit from collaboration and 
interaction with Ross Fadely (Insight), ChangHoon Hahn (NYU), Andrew Hearin (Yale), Dan Foreman-Mackey (UW), 
Kilian Walsh (NYU), Paco Kitauro (IAC), Yu Feng (Berkeley), Albert Chuang (Leibnitz), Boris Leistedt (NYU), 
and Alex Malz (NYU).

I would like to acknowledge the fantastic computational support that I received at NYU. 
I am grateful to Shenglong Wang (the administrator of NYU HPC) and Mulin Ding (the administrator 
of CCPP computational facilities) for their persistent support and phenomenal work ethics. 

I truely enjoyed the lively and intellectually stimulating atmosphere at CCPP.
I also want to thank all my great friends in graduate school who helped create a 
supportive and joyful environment.  

I thank my family who have always supported me throughout my entire life.

The work that led to Chapter 1 of this dissertation was partially supported by the NSF (grants IIS-1124794 and AST-1517237), 
NASA (grant NNX12AI50G), and the Moore-Sloan Data Science Environment at NYU. 
I thank Jo Bovy and Alex Malz for discussions related to this work. 
I am also grateful to Dustin Lang and Alex Malz for reading and making comments on draft. 
I am also grateful to Jay~Anderson, Dan~Foreman-Mackey, Michael~Blanton, and Alex~Malz for discussions related to the second chapter 
of thesis.

I thank Jessie Cisewsky for reading and making valuable comments on the draft of the paper that the third chapter of this dissertation 
is based on. I would also like to thank Michael R. Blanton, Jeremy R. Tinker, Uros Seljak, Layne Price, 
Boris Leidstadt, Alex Malz, Patrick McDonald, and Dan Foreman-Mackey for productive 
and insightful discussions. This work was supported by NSF grant AST-1517237. Computations 
were performed using computational resources at NYU-HPC. I thank Shenglong Wang, the 
administrator of NYU-HPC computational facility, for his consistent and continuous support 
throughout the development of this project. I would like to thank the organizers of 
the AstroHackWeek 2015 workshop (\url{http://astrohackweek.org/2015/}), 
since the direction and the scope of this investigation was ---to some degree--- initiated 
through discussions in this workshop. Throughout this investigation, I made use of 
publicly available software packages $\mathtt{emcee}$ and $\mathtt{abcpmc}$. 
I have also used the publicly available python implementation of the FoF algorithm 
$\mathtt{pyfof}$ (\url{https://github.com/simongibbons/pyfof}).

I am grateful to David~W.~Hogg, Alex~I.~Malz, Andrew~Hearin, Andrew~Zentner, Chia-Hsun~Chuang, Michael~R.~Blanton, and Kilian~Walsh for discussions related to Chapter 4. This work was supported by the NSF grant AST-1517237. All the computations in this work were carried out in New York University High Performance Computing Mercer facility. 
The SDSS clustering measurements and the covariance matrices used in Chapter 4 are available at \url{http://sdss4.shao.ac.cn/guoh/files/wpxi_measurements_Guo15.tar.gz}. Description of the $\mathtt{SMDP}$ and the $\mathtt{BolshoiP}$ halo catalogs used in this investigation can be found at \url{https://www.cosmosim.org/cms/simulations}. The RockStar halo catalogs of the $\mathtt{SMDP}$ and the $\mathtt{BolshoiP}$ simulations are publicly available at \url{http://yun.ucsc.edu/sims/SMDPL/hlists/index.html} and \url{http://yun.ucsc.edu/sims/Bolshoi_Planck/hlists/index.html} respectively. We thank Peter Behroozi for making the halo catalogs publicly available. In this work we have made use of the publicly available codes: corner (\citealt{corner}), emcee (\citealt{emcee}), halotools (\citealt{halotools}), Corrfunc (\citealt{corrfunc}), and changtools (\url{https://github.com/changhoonhahn/ChangTools}). The abundance matching mock catalogs used in this study are available at (\url{http://logrus.uchicago.edu/~aphearin/}).  

I am grateful to David~W.~Hogg, Uros~Seljak, Martin~White, Emanuele~Castorina, Jeremy~L.~Tinker, and Michael~Blanton for discussions related to Chapter 5.
I am particularly thankful to David~W.~Hogg for his continuous support during the completion of this work. I also thank David~W.~Hogg and Alex~I.~Malz for reading and commenting on the manuscript. This work was supported by the NSF grant AST-1517237. Most of the computations in this work were carried out in the New York University High Performance Computing Mercer facility. We thank Shenglong~Wang, the administrator of the NYU HPC center, for his consistent support throughout the completion of this study.
 
The CosmoSim database used in Chapter 4 and 5 is a service by the Leibniz-Institute for Astrophysics Potsdam (AIP). The MultiDark database was developed in cooperation with the Spanish MultiDark Consolider Project CSD2009-00064. The authors gratefully acknowledge the Gauss Centre for Supercomputing e.V. (www.gauss-centre.eu) and the Partnership for Advanced Supercomputing in Europe (PRACE, www.prace-ri.eu) for funding the MultiDark simulation project by providing computing time on the GCS Supercomputer SuperMUC at Leibniz Supercomputing Centre (LRZ, www.lrz.de).

